---
title: "ART FROM CODE"
author: "Danielle Navarro"
execute:
  echo: true
  fig-width: 6
  fig-height: 6
format: 
  revealjs:
    theme: dark
    footer: "[art-from-code.netlify.app](https://art-from-code.netlify.app)"
---

```{r}
#| include: false

# can't seem to make this work in the yaml so...
knitr::opts_chunk$set(dev.args = list(bg="#191919"))

# save the built-in output hook
hook_output <- knitr::knit_hooks$get("output")

# set a new output hook to truncate text output
knitr::knit_hooks$set(output = function(x, options) {
  if (!is.null(n <- options$out.lines)) {
    x <- xfun::split_lines(x)
    if (length(x) > n) {
      # truncate the output
      x <- c(head(x, n), "...\n")
    }
    x <- paste(x, collapse = "\n")
  }
  hook_output(x, options)
})
```

```{r packages, include=FALSE}
library(ggplot2)
library(tibble)
library(dplyr)
library(ggthemes)
library(scales)
```

## WiFi
	
- Username: conf22
- Password: together!

## Facilities
	
- Gender neutral bathroom: 
    - river side of Potomac Ballroom
- Meditation room: 
    - National Harbor 9. (8am - 5pm, Mon - Thurs)
- Lactation room: 
    - Potomac Dressing Room. (8am - 5pm, Mon - Thurs) 
    - The hotel also has a dedicated room behind reception

## Code of conduct

- The Code of Conduct and COVID policies can be found at [www.rstudio.com/conference/2022/2022-conf-code-of-conduct](https://www.rstudio.com/conference/2022/2022-conf-code-of-conduct/). Please review them carefully
- You can report Code of Conduct violations in person, by email, or by phone. Please see the policy linked above for contact information

## Mask policy

- RStudio requires that you wear a mask that fully covers your mouth and nose at all times in all public spaces 
- We strongly recommend that you use a correctly fitted N95, KN95, or similar particulate filtering mask
- We will have a limited supply available upon request

## Photography

- Red lanyard = Does not wish to be photographed 
- Please note lanyard colors before taking a photo
- Respect their choices

## Workshop housekeeping

- Written tutorial, data sets, exercises
    - Slides: Intentionally minimal! We'll work from the notes :)
    - Website: [art-from-code.netlify.app](https://art-from-code.netlify.app)
    - GitHub: [github.com/rstudio-conf-2022/art-from-code](https://github.com/rstudio-conf-2022/art-from-code)
- Instructors: 
    - Danielle Navarro
    - Meghan Harris
    - Jaclyn Janis 

## Schedule for day 1

| Time          | Activity                                                                              |
| :------------ | :------------------------------------------------------------------------------------ |
| 09:00 - 10:30 | Session 1: [Get started](https://art-from-code.netlify.com/day-1/session-1)           |
| 10:30 - 11:00 | *Coffee break*                                                                        |
| 11:00 - 12:30 | Session 2: [Spatial noise tricks](https://art-from-code.netlify.com/day-1/session-2)  |
| 12:30 - 13:30 | *Lunch break*                                                                         |
| 13:30 - 15:00 | Session 3: [Polygon tricks](https://art-from-code.netlify.com/day-1/session-3)        |
| 15:00 - 15:30 | *Coffee break*                                                                        |
| 15:30 - 17:00 | Session 4: [Shading tricks](https://art-from-code.netlify.com/day-1/session-4)        |

## Schedule, Day 2

| Time          | Activity                                                                                  |
| :------------ | :---------------------------------------------------------------------------------------- |
| 09:00 - 10:30 | Session 1: [Iterated function systems](https://art-from-code.netlify.com/day-2/session-1) |
| 10:30 - 11:00 | *Coffee break*                                                                            |
| 11:00 - 12:30 | Session 2: [Tiles and tesselations](https://art-from-code.netlify.com/day-2/session-2)    |
| 12:30 - 13:30 | *Lunch break*                                                                             |
| 13:30 - 15:00 | Session 3: [Pixel shaders](https://art-from-code.netlify.com/day-2/session-3)             |
| 15:00 - 15:30 | *Coffee break*                                                                            |
| 15:30 - 17:00 | Session 4: [Wrap up](https://art-from-code.netlify.com/day-2/session-4)                   |

## Installing packages

TBA

## Downloading materials

TBA

# [Get started](https://art-from-code.netlify.com/day-1/session-1)

Session 1: Hands on introduction to generative art. Make your own pieces using ggplot2 and dplyr

# [Spatial noise tricks](https://art-from-code.netlify.com/day-1/session-2)

Session 2: Multidimensional noise generators. Flow fields, fractals, and more using the ambient package

# [Polygon tricks](https://art-from-code.netlify.com/day-1/session-3)

Session 3: Create textures with recursive polygon deformation, draw wonky hearts as animated gifs, and convert boring lines to textured strokes

# [Shading tricks](https://art-from-code.netlify.com/day-1/session-4) 

Session 4: Use rayshader to add depth to 2D plots, and generate 3D graphics from generative art

# Welcome to Day 2

# [Iterated function systems](https://art-from-code.netlify.com/day-2/session-1)

Session 1: Embrace the chaotic joy of the fractal flame and render quickly with Rcpp. This one is a bit technical

# [Tiles and tesselations](https://art-from-code.netlify.com/day-2/session-2) 

Session 2: Make lovely things by subdividing a rectangle, explore Voronoi tesselations, and have a passing encounter with Truchet tiles

# [Pixel shaders](https://art-from-code.netlify.com/day-2/session-3) 

Session 3: Manipulate ggplot objects as if they were raster graphics using ggfx: learn to dither, mask, displace, blend, and more. The flametree package makes as special guest appearance

# [Wrap up](https://art-from-code.netlify.com/day-2/session-4)

Session 4: In which we end things by making things up as we go along, because this is, after all, an art workshop!


