---
title: "SPATIAL TRICKS WITH AMBIENT"
---


```{r package-load}
#| message: false
library(dplyr)
library(purrr)
library(tibble)
library(ggplot2)
library(ggthemes)
library(ambient)
```

```{r seed, include=FALSE}
set.seed(1)
```

## Sampling spatial patterns

Generative art relies on having access to a source of randomness, and using that randomness to construct patterned objects. In the last session I wrote a simple function to generate random palettes, for instance:

```{r sample-canva}
sample_canva <- function(seed = NULL) {
  if(!is.null(seed)) set.seed(seed)
  sample(ggthemes::canva_palettes, 1)[[1]]
}
```

Every time I call this function (assuming I don't set the `seed` argument), R uses the pseudorandom number generator to select a palette of four colours:

```{r sample-canva-examples, results='hold'}
sample_canva()
sample_canva()
sample_canva()
```

Each of these colours is itself an object defined in a three dimensional space (the hex codes refer to co-ordinates in RGB space), and when I sample a palette of four colours what I'm really doing is constructing a random object with 12 components. 

We can take this idea further. The `sample_data()` function I wrote in the last session creates random tibbles according to some simple rules, and those tibbles are structured objects too. Admittedly the structure to those objects isn't very complicated, because there's no pattern to numbers in the generated table, but there's nothing stopping us from writing a function that randomly generates tabular data structures that have patterns in them, right? For instance, I could do this:

```{r, sample_cross_matrix}
sample_cross_matrix <- function(n = 10, seed = NULL) {
  if(!is.null(seed)) set.seed(seed)
  mat <- matrix(data = 0, nrow = n, ncol = n)
  mat[sample(n, 1), ] <- 1
  mat[, sample(n, 1)] <- 1
  return(mat)
}

sample_cross_matrix()
```

Again, this isn't the most complicated example, but it's an illustration of the idea that we can write functions to sample random spatial patterns:

```{r random-crosses, dev.args=list(bg="#222222")}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| column: screen-inset
#| layout-ncol: 3
#| cache: true
image(sample_cross_matrix(n = 50), axes = FALSE, useRaster = TRUE)
image(sample_cross_matrix(n = 50), axes = FALSE, useRaster = TRUE)
image(sample_cross_matrix(n = 50), axes = FALSE, useRaster = TRUE)
```

I'll concede that a generative art system that draws a cross at a random location in the image isn't the most exciting or innovative thing I've ever written, but the core idea is clear I hope. And you would probably be unsurprised to learn that there are a number of more sophisticated tools you can use to generate random spatial patterns. 

In this session I'll introduce the [ambient package](https://ambient.data-imaginist.com/), developed by [Thomas Lin Pedersen](https://www.data-imaginist.com/), which supplies R bindings to a C++ library called FastNoise.

## Our first ambient artwork

The first step in creating art using the `ambient` package is to define the "canvas", a spatial grid of x and y co-ordinates in which values will be stored. I find it helpful to imagine my canvas as the unit square: the smallest value is 0 and the largest value is 1. If I want an 800x800 grid, I use the `seq()` function to define a length-800 sequence of evenly-spaced numbers starting at 0 and ending at 1:

```{r define-coords}
x_coords <- seq(from = 0, to = 1, length.out = 800)
y_coords <- seq(from = 0, to = 1, length.out = 800)
```

The `canvas` that I'll paint on will be a data frame consisting of all possible combinations of `x_coords` and `y_coords`. In base R we can create an object like this using the `expand.grid()` function, and there's a tidy equivalent called `expand_grid()` in the `tidyr` package. However, when working with the `ambient` package I prefer to use the `long_grid()` function that it supplies:

```{r define-canvas}
#| cache: true
canvas <- long_grid(x = x_coords, y = y_coords) 
canvas
```

The reason I use `long_grid()` rather than one of the more familiar versions is that under the hood Thomas has supplied some optimisations that make these objects more efficient for generative art purposes. Among other things, you can easily convert them to arrays, matrices, and raster objects that respect the implied spatial grid, which makes it a lot easier to render images from these objects. But let's not dive too deep into the details right now!

Now that we have a "canvas" we'll want to add some "paint", and to apply that paint we'll need to select a "brush". In this context, the brush is a spatial pattern generator of some kind. The `ambient` package includes many such generator functions. Some generate very regular patterns:

- `gen_waves()` generates smooth concentric wave-like patterns
- `gen_spheres()` generates concentric circles
- `gen_checkerboard()` generates grids of squares in a checkerboard pattern

Others generate very irregular patterns:

- `gen_white()` generates white noise: equal intensity at every spatial frequency

The most interesting generators tend to be those that create patterns that have some structure but are still quite unpredictable:

- `gen_perlin()` and `gen_simplex()` generate random "wavy" patterns
- `gen_worley()` generates random "cellular" patterns

We'll see some examples of those later! For now let's use `gen_perlin()` as our brush! Like all the pattern generators, the `gen_perlin()` function takes coordinate values as input. At a minimum it expects an `x` co-ordinate, but you can also supply `y` and `z` values if you like. You can also specify the `frequency` parameter, which sets the scale for the output: high frequency patterns will vary quickly as the co-ordinates change, low-frequency patterns will vary slowly. Compare these outputs for instance:

```{r perlin-examples, results='hold'}
gen_perlin(x = 1:5, y = 1, frequency = .001, seed = 1)
gen_perlin(x = 1:5, y = 1, frequency = .5, seed = 1)
```

Both versions show variability, but the scale is quite different! As an aside, notice that `gen_perlin()` also allows you to specify the `seed` used to generate the pattern, similar to the way I did earlier when writing `sample_canva()` to generate random palettes. 

Now that we have a sense for what the `gen_perlin()` function does let's use it to add a new column to our `canvas`. I have `dplyr` loaded so I'll use `mutate()` to do this:

```{r add-perlin-noise}
#| cache: true
canvas <- canvas |> 
  mutate(paint = gen_perlin(x, y, frequency = 10, seed = 1234))
canvas
```

Now that I have this column, I can use it to control the `fill` aesthetic in a ggplot. The co-ordinate values `x` and `y` specify a two dimensional grid, which means I can use `geom_raster()` here to create my artwork:

```{r}
#| cache: true
art <- ggplot(canvas, aes(x, y, fill = paint)) + 
  geom_raster(show.legend = FALSE) 
```

To see what our Perlin art looks like, I'll build the plot in three different ways. First poorly with no customisation of the ggplot theme and scales, then a little nicer by removing unneeded details, then finally with a little flair:

```{r simple-perlin-art}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| column: screen-inset
#| layout-nrow: 1
#| cache: true
art
art + 
  theme_void() +
  coord_equal()
art + 
  theme_void() +
  coord_equal() +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_gradientn(colours = sample_canva())
```

Not bad. A little blurry looking, but it's a nice place to start!



::: {.callout-important icon=false #exercise-first-ambient}
## Exercise

Try it yourself! In the materials folder there is a file called `first-ambient-art.R` that reproduces the code above. Try playing around with it to see what kind of output you can create by changing the values fed to `gen_perlin()`, or by trying other generator functions!

:::



## Our first system

The next step in the process is to start thinking about what aspects to the art should be variable, what aspects should be fixed, and use those insights to formalise this as a function. Some things that won't change:

- The art will always be a square grid rendered with `geom_raster()`
- The spatial pattern will always come from one of the `ambient::gen_*()` functions
- We'll always remove extraneous elements from the art using `theme_void()` etc

These aspects to the art will be codified in the function body. There are some things that we might like to vary, however, and those will become arguments to the function:

- The colour `palette` could vary from piece to piece
- The underlying `generator` might be different in each piece
- The spatial `frequency` could be different in each piece
- The number of `pixels` in the grid could vary
- The random number generator `seed` could vary 


```{r simple-art-function}
#| cache: true
make_noise_art <- function(
    generator = gen_perlin, 
    frequency = 10, 
    seed = 1234,
    pixels = 2000,
    palette = c("#e5ddc8", "#01949a", "#004369", "#db1f48"), 
    ...
) {
  
  # define the grid
  canvas <- long_grid(
    x = seq(from = 0, to = 1, length.out = pixels),
    y = seq(from = 0, to = 1, length.out = pixels)
  ) 
  
  # use the generator to add paint
  canvas <- canvas |>
    mutate(
      paint = generator(
        x, y, 
        frequency = frequency, 
        seed = seed, 
        ...
      )
    )
  
  # use ggplot2 to draw the picture
  art <- canvas |> 
    ggplot(aes(x, y, fill = paint)) + 
    geom_raster(show.legend = FALSE) +
    theme_void() +
    coord_equal() +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    scale_fill_gradientn(colours = palette)
  
  return(art)
}
```

Let's take a lot the effect of each of these arguments. Varying `seed` changes the spatial pattern depicted in each piece, but it doesn't change it in any systematic way:

```{r example-art-seed}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| column: screen-inset
#| layout-nrow: 1
#| cache: true

make_noise_art(seed = 1234)
make_noise_art(seed = 1001)
make_noise_art(seed = 9999)
```

In contrast, when I change the `frequency` argument I get systematic variation. The granularity of the spatial pattern changes in predictable ways as the frequency changes:

```{r example-art-frequency}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| column: screen-inset
#| layout-nrow: 1
#| cache: true

make_noise_art(frequency = 10)
make_noise_art(frequency = 20)
make_noise_art(frequency = 90)
```


Here's what happens when I vary `palette`. In the first example I've created a greyscale image by specifying a palette that runs from `"white"` to `"black"`. The other two use palettes output by `sample_canva()`:

```{r example-art-palette}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| column: screen-inset
#| layout-nrow: 1
#| cache: true
make_noise_art(palette = c("white", "black"))
make_noise_art(palette = sample_canva(seed = 123))
make_noise_art(palette = sample_canva(seed = 456))
```


Finally, we can vary the `generator` function. It probably will come as no surprise to discover that varying the generator has some wild effects. The output of a checkerboard pattern generator is fundamentally different to the output of a Worley noise generator, which in turn is very distinct from Perlin noise. As you become more familiar with using `ambient` you'll start getting a sense of what each of these generators produce, and develop your own preferences for how to use them. For now, it's enough to note that because the `gen_*()` functions all adopt (roughly) the same API, our `make_noise_art()` function works perfectly well when we swap out one for another:


```{r example-art-generator}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| column: screen-inset
#| layout-nrow: 1
#| cache: true
make_noise_art(generator = gen_perlin)
make_noise_art(generator = gen_worley)
make_noise_art(generator = gen_waves) 
```


::: {.callout-important icon=false #exercise-first-ambient}
## Exercises

1. In the materials folder there is a file called `make-noise-art.R` that includes the `make_noise_art()` function. Unlike the code I've shown here, the version in the script writes the output to a file (located at `output/noise-art.png`). Try playing around with the inputs to `make_noise_art()` to see what outputs you can create.

2. At the moment, the script is set up so that the output is always written to the same file, `output/noise-art.png`. When you create your own generative art systems you will want to ensure that each unique output is written to a file with a unique filename, and that this filename should (ideally!) allow you to work out what inputs were used to create the piece. How would you write code to do this?

:::


## Why dplyr is a girls best friend

As you can see from the output we've created so far, spatial noise patterns can be quite pretty even without any special artistic intervention. Our `make_noise_art()` function isn't complicated: it takes the output from a generator function like `gen_perlin()` and plots it as a raster object. It doesn't manipulate or modify that output in any way. However, there's nothing preventing us from doing precisely that if that's what we want to do. To simplify the later code, let's create a `blank_canvas` object that we can reuse as the starting point for our later pieces:

```{r blank_canvas}
blank_canvas <- long_grid(
  x = seq(from = 0, to = 1, length.out = 2000),
  y = seq(from = 0, to = 1, length.out = 2000)
) 
```

Now, let's imagine that we've used some `ambient` magic to add a column called `paint` to our canvas. Here's a plotting function that we can use that plots this as a raster object, just like we've been doing in the previous pieces (it optionally takes a `palette` too):

```{r dplyr-art-setup}
plot_painted_canvas <- function(canvas, palette = NULL) {
  if(is.null(palette)) {
    palette <- c("#e5ddc8","#01949a","#004369","#db1f48")
  }
  canvas |> 
    ggplot(aes(x, y, fill = paint)) + 
    geom_raster(show.legend = FALSE) +
    theme_void() +
    coord_equal() +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    scale_fill_gradientn(colours = palette)
}
```

Now that we have these in place, we can recreate one of our earlier pieces by wrting it as a `dplyr` pipeline:

```{r dplyr-recreation}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| fig-align: center
#| cache: true
blank_canvas |>
  mutate(paint = gen_perlin(x, y, frequency = 90, seed = 1234)) |>
  plot_painted_canvas()
```

However, the mere fact that we can rewrite our art code like this opens up the possibility of using the `dplyr` data manipulation grammar in a more sophisticated way. Here's an example that creates three different spatial patterns and then adds them together:

```{r dplyr-art-example-1}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| fig-align: center
#| cache: true
blank_canvas |> 
  mutate(
    lf_noise = gen_simplex(x, y, frequency = 1, seed = 1234),
    mf_noise = gen_simplex(x, y, frequency = 20, seed = 1234),
    hf_noise = gen_simplex(x, y, frequency = 99, seed = 1234),
    paint = lf_noise + mf_noise + hf_noise
  ) |>
  plot_painted_canvas()
```

Recall that our `plot_painted_canvas()` function uses the `x` and `y` columns to define the grid, and the `paint` column to define the to-be-plotted values. The `lf_noise`, `mf_noise`, and `hf_noise` columns are ignored. They're intermediate steps, components that get mixed together when we define the `paint` column!

In the previous example I created the `paint` column by adding three columns together, but there is nothing preventing me from defining more elaborate mixing rules. In the example below, for example, I've generated a fourth spatially-varying pattern and used as a "gating" mechanism. So now what we have is a situation where the `lf_noise`, `mf_noise`, and `hf_noise` patterns are mixed together in a spatially inhomogeneous way that depends on the value of the `gate` column:

```{r dplyr-art-example-2}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| fig-align: center
#| cache: true
blank_canvas |> 
  mutate(
    lf_noise = gen_simplex(x, y, frequency = 1),
    mf_noise = gen_simplex(x, y, frequency = 20),
    hf_noise = gen_simplex(x, y, frequency = 99),
    gate = gen_spheres(x, y, frequency = 10) |> normalise(),
    paint = lf_noise +
      (1 + mf_noise) * (gate >= .1 & gate < .6) +
      (1 + hf_noise) * (gate >= .05)
  ) |>
  plot_painted_canvas(palette = sample_canva(seed = 2))
```

The `normalise()` function in this code is supplied by the `ambient` package and in this context all I'm doing with it is ensuring that the output of the `gen_spheres()` generator is rescaled to lie between 0 and 1. 

The same basic idea can be used to produce some quite striking pieces when we apply a fancier generator to construct the spatial `gate` pattern:

```{r dplyr-art-example-3}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| fig-align: center
#| cache: true
blank_canvas |> 
  mutate(
    lf_noise = gen_simplex(x, y, frequency = 1),
    mf_noise = gen_simplex(x, y, frequency = 20),
    hf_noise = gen_simplex(x, y, frequency = 99),
    gate = gen_simplex(x, y, frequency = 10) |> normalise(),
    paint = lf_noise +
      (2 + mf_noise) * (gate >= .2 & gate < .8) +
      (2 + hf_noise) * (gate >= .1)
  ) |>
  plot_painted_canvas(palette = sample_canva(seed = 3))
```



::: {.callout-important icon=false #exercise-dplyr-ambient}
## Exercise

Try it yourself! In the materials folder there is a script called `dplyr-ambient.R` that defines `blank_canvas` and `plot_painted_canvas()` for you. At the bottom of the file there is space for you to add to the blank canvas. Try using dplyr and ambient together to create a spatial noise pattern of your own.

:::



## Fractals

You can also use the `ambient` package to create fractal patterns. The function that controls this is called `fracture()` and it's easiest to demonstrate if we start with something simple. Suppose we have a "generator" function `gen_sin()` that generates sinusoidal patterns with a particular frequency. The code for this function is very simple:

```{r gen-sin}
gen_sin <- function(x, frequency, ...) {
  sin(x * frequency)
}
```

To create a fractal pattern based on this generator, we repeatedly apply this function to the input at different values of `frequency`. The outputs of repeated applications are combined together using a rule prescribed by a `fractal` function. This combination function doesn't have to be very complicated: it might just be a linear combination! As an example, one of the fractal functions provided by `ambient` is `fbm()`, which stands for "fractional Brownian motion". When this is used as the combination rule, the results are added together with increasing frequencies and decreasing strength. The code for `fbm()` is this:

```{r}
fbm <- function(base, new, strength, ...) {
  base + new * strength
}
```

A `base` pattern is added to a `new` pattern, weighted by some `strength`. That's all it does!

If we wanted to create a fractal based on the `gen_sin()` generator, using `fbm()` as our fractal function, this is the code we would use:

```{r}
fracture(
  x = 1:20, 
  noise = gen_sin, 
  fractal = fbm, 
  octaves = 8
)
```

In this code, `octaves = 8` specifies the number of times to apply the generator and fractal function. In essence it is the number of iterations over which we run the algorithm. It's easiest to see what this looks like if we gradually increase the number of iterations and plot the results:


```{r fracture-api}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| column: screen-inset
#| layout-nrow: 1
#| cache: true
dat <- tibble(
  x = seq(0, 10, length.out = 1000), 
  y1 = fracture(x = x, noise = gen_sin, fractal = fbm, octaves = 1),
  y2 = fracture(x = x, noise = gen_sin, fractal = fbm, octaves = 2),
  y8 = fracture(x = x, noise = gen_sin, fractal = fbm, octaves = 8),
  y20 = fracture(x = x, noise = gen_sin, fractal = fbm, octaves = 20)
) 

ggplot(dat) + geom_path(aes(x, y1)) + ggtitle("One iteration")
ggplot(dat) + geom_path(aes(x, y2)) + ggtitle("Two iterations")
ggplot(dat) + geom_path(aes(x, y8)) + ggtitle("Eight iterations")
ggplot(dat) + geom_path(aes(x, y20)) + ggtitle("Twenty iterations")
```

As the number of octaves increases the plots become more and more detailed. In this case we can't visually discriminate between 8 and 20 octaves because the differences are too fine-grained to be visible. That's quite typical because -- unless you modify the `gain` and `frequency` functions used by `fracture()` -- each successive iteration (or octave) will be calculated at double the frequency of the previous one (leading to finer-grained changes) and with half the strength (less weight is given to later octaves). You can modify this if you want to. For example:

```{r custom-fracture}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| column: screen-inset
#| layout-nrow: 1
#| cache: true

custom_fracture <- function(x, ...) {
  fracture(
    gain = function(strength) {strength * .8},
    frequency = function(frequency) {frequency * 1.3},
    noise = gen_sin, 
    fractal = fbm,
    x = x,
    ...
  )
}

dat <- tibble(
  x = seq(0, 10, length.out = 1000), 
  y1 = custom_fracture(x, octaves = 1),
  y2 = custom_fracture(x, octaves = 2),
  y8 = custom_fracture(x, octaves = 8),
  y20 = custom_fracture(x, octaves = 20)
) 

ggplot(dat) + geom_path(aes(x, y1)) + ggtitle("One iteration")
ggplot(dat) + geom_path(aes(x, y2)) + ggtitle("Two iterations")
ggplot(dat) + geom_path(aes(x, y8)) + ggtitle("Eight iterations")
ggplot(dat) + geom_path(aes(x, y20)) + ggtitle("Twenty iterations")
```

Hopefully you get the basic idea. 

In any case, let's take this same concept and start using it in conjunction with the spatial noise generators supplied by `ambient`. To keep things simple and avoid the need to write plotting code over and over, let's define a `fractal_art()` function like this:

```{r fractal-art-function}
#| cache: true
fractal_art <- function(fractal, generator, palette = NULL, ...) {
  blank_canvas |>
    mutate(
      paint = fracture(
        noise = generator,
        fractal = fractal,
        x = x, 
        y = y, 
        ...
      )
    ) |>
    plot_painted_canvas(palette = palette)
}
```

Here's what happens when we use `gen_checkerboard()` as our spatial pattern generator, and `fbm()` as our fractal function:

```{r fractal-art-checker}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| column: screen-inset
#| layout-nrow: 1
#| cache: true
fractal_art(fbm, gen_checkerboard, seed = 1, octaves = 1)
fractal_art(fbm, gen_checkerboard, seed = 1, octaves = 2)
fractal_art(fbm, gen_checkerboard, seed = 1, octaves = 20)
```

It has the same "feel" as the sinusoidal fractals we were working with earlier: as we increase the number of octaves the output contains more copies of the "checker board" pattern, each one depicted on a smaller scale than the last one. The same idea applies to the `gen_waves()` generator:

```{r fractal-art-spheres}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| column: screen-inset
#| layout-nrow: 1
#| cache: true
fractal_art(fbm, gen_waves, seed = 1, octaves = 1)
fractal_art(fbm, gen_waves, seed = 1, octaves = 2)
fractal_art(fbm, gen_waves, seed = 1, octaves = 20)
```

By the time we reach 20 octaves, the image has a quite intricate pattern of concentric rings. It's quite pretty, but `gen_checkerboard()` and `gen_waves()` are both very simple generator functions. What happens when our generator is a more elaborate multidimensional noise generator like `gen_simplex()`? Simplex noise looks like this before we apply any fractal function to it:

```{r base-simplex}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| fig-align: center
#| cache: true
blank_canvas |>
  mutate(paint = gen_simplex(x, y, seed = 2)) |>
  plot_painted_canvas()
```

Here's what happens when we combine `gen_simplex()` with the `fbm()` fractal function:

```{r fractal-art-octaves}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| column: screen-inset
#| layout-nrow: 1
#| cache: true
fractal_art(fbm, gen_simplex, seed = 2, octaves = 1)
fractal_art(fbm, gen_simplex, seed = 2, octaves = 2)
fractal_art(fbm, gen_simplex, seed = 2, octaves = 20)
```

The result, once we reach 20 octaves, is quite intricate. 

Changing the fractal function has a substantial effect on the output. So far all the fractals I've created have used `fbm()` as the fractal function, but there's nothing stopping you from writing your own or using one of the other functions supplied by `ambient`. For example, the `ridged()` fractal function produces some very lovely patterns:


```{r fractal-art-ridged-octaves}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| column: screen-inset
#| layout-nrow: 1
#| cache: true
fractal_art(ridged, gen_simplex, seed = 2, octaves = 1)
fractal_art(ridged, gen_simplex, seed = 2, octaves = 2)
fractal_art(ridged, gen_simplex, seed = 2, octaves = 20)
```

It's also possible to get nice effects by modifying the `gain` and `frequency` functions. For example, here's an example where the strength of each successive iteration of `ridged()` noise diminishes to 80% of of the strength of the previous iteration:


```{r fractal-art-ridged-gain}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| column: screen-inset
#| layout-nrow: 1
#| cache: true
gf <- function(x) x * .8
fractal_art(ridged, gen_simplex, seed = 2, octaves = 1, gain = gf)
fractal_art(ridged, gen_simplex, seed = 2, octaves = 2, gain = gf)
fractal_art(ridged, gen_simplex, seed = 2, octaves = 20, gain = gf)
```

Worley noise is an interesting case. The behaviour of `gen_worley()` is to carve the image up in to distinct cells, and colour each pixel in the image based on the cells they belong to. It's closely related to Voronoi tesselation, a technique I'll talk about in a later session. In the simplest case, all pixels in a particular cell are assigned the same colour. So a very basic Worley noise pattern might look like this:

```{r base-worley}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| fig-align: center
#| column: screen-inset
#| layout-nrow: 1
#| cache: true
blank_canvas |>
  mutate(paint = gen_worley(x, y, seed = 6)) |>
  plot_painted_canvas()
```

When we create fractals using this kind of generator, there's a tendency to end up with "kaleidoscopic" looking patterns. Here's an example using the `billow()` fractal function:

```{r fractal-art-worley-octaves}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| column: screen-inset
#| layout-nrow: 1
#| cache: true
fractal_art(billow, gen_worley, seed = 6, octaves = 1)
fractal_art(billow, gen_worley, seed = 6, octaves = 3)
fractal_art(billow, gen_worley, seed = 6, octaves = 8)
```

However, `gen_worley()` allows also allows you to colour the pixels in different ways. For example, if I set `value = "distance"`, each pixel will be coloured as a function of how distant it is from the centroid of the cell it belongs to. A basic pattern looks like this:

```{r base-worley-distance}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| fig-align: center
#| column: screen-inset
#| layout-nrow: 1
#| cache: true
blank_canvas |>
  mutate(paint = gen_worley(x, y, seed = 6, value = "distance")) |>
  plot_painted_canvas()
```

Fractals created using this method look like this:

```{r fractal-art-worley-distance}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| column: screen-inset
#| layout-nrow: 1
#| cache: true
fractal_art(billow, gen_worley, seed = 6, octaves = 1, value = "distance")
fractal_art(billow, gen_worley, seed = 6, octaves = 3, value = "distance")
fractal_art(billow, gen_worley, seed = 6, octaves = 8, value = "distance")
```

There's nothing stopping you from writing your own generator function either. At the start of this section that's exactly what I did in one dimension with `gen_sin()`. As a two dimensional example, let's suppose I wanted to create a variation of Worley noise that mixes both the `"cell"` colouring and the `"distance"` colouring. Here's a generator function that does precisely that:

```{r base-worley-mixed}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| fig-align: center
#| column: screen-inset
#| layout-nrow: 1
#| cache: true
gen_scope <- function(x, y, ...) {
  worley_cell <-gen_worley(x, y, value = "cell", ...)
  worley_dist <-gen_worley(x, y, value = "distance", ...)
  return(normalise(worley_cell) + 5 * normalise(worley_dist))
}

pal <- sample_canva(seed = 2)

blank_canvas |>
  mutate(paint = gen_scope(x, y, seed = 9)) |>
  plot_painted_canvas(palette = pal)
```

I can now use my `gen_scope()` function as the generator for my fractal:

```{r fractal-art-scope}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| column: screen-inset
#| layout-nrow: 1
#| cache: true
fractal_art(billow, gen_scope, palette = pal, seed = 9, octaves = 1)
fractal_art(billow, gen_scope, palette = pal, seed = 9, octaves = 2)
fractal_art(billow, gen_scope, palette = pal, seed = 9, octaves = 8)
```

I can make the generator as elaborate as I like. The `gen_gate()` function below uses a "gating" mechanism just like the example I used earlier:

```{r fractal-art-gated}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| column: screen-inset
#| layout-nrow: 1
#| cache: true
gen_gate <- function(x, y, frequency, ...) {
  lf <- gen_simplex(x, y, frequency = frequency, ...)
  mf <- gen_simplex(x, y, frequency = frequency * 20, ...)
  hf <- gen_simplex(x, y, frequency = frequency * 99, ...)
  gate <- gen_simplex(x, y, frequency = frequency * 10, ...) 
  gate <- normalise(gate)
  paint <- lf + 
    (mf + 2) * (gate >= .2 & gate < .8) + 
    (hf + 2) * (gate >= .1)
  return(paint)
}

pal <- sample_canva(seed = 3)

fractal_art(billow, gen_gate, palette = pal, seed = 9, octaves = 1)
fractal_art(billow, gen_gate, palette = pal, seed = 9, octaves = 2)
fractal_art(billow, gen_gate, palette = pal, seed = 9, octaves = 20)
```


::: {.callout-important icon=false #exercise-fractal-art}
## Exercises

The `fractal_art.R` script in the materials folder contains all the setup you need to play around with the `fractal_art()` function. Try using it to explore the possibilities! There are a lot of possibilities in fractals. Here are a few ideas to get you started:

1. The easiest way to play around with fractals is to modify the basic arguments. Try changing the `generator`, `fractal`, `freq_init` (i.e., frequency value for the first octave), `octaves`, `seed`, and `palette` to make a piece you really like. 

2. A fun inversion: use those same arguments to create something that you find incredibly ugly!

3. The `fractal_art()` function is written flexibly enough that you can pass your own gain functions and frequency functions. There's an example of this above. Try writing your own functions to modify the `gain` and `frequency` rules that apply to the fractal.

4. Using the `gen_scope()` and `gen_gate()` examples to motivate you, write your own generator function. See what effect that has.

:::


## Curl (of a spatial) noise (pattern)

The last topic to talk about in regards to the `ambient` package is curl noise. The concept comes from vector calculus, I'm afraid, but fortunately for us we don't actually need to care. To quote from Wikipedia, the [curl](https://en.wikipedia.org/wiki/Curl_(mathematics)) is

> a vector operator that describes the infinitesimal circulation of a vector field in three-dimensional Euclidean space. The curl at a point in the field is represented by a vector whose length and direction denote the magnitude and axis of the maximum circulation. The curl of a field is formally defined as the circulation density at each point of the field. 

Exciting stuff. But what does it *mean*? Well, let's suppose I have a vector field and... 

... wait, what?   

Okay, let's take a step back. Suppose I have a very small grid of points:

```{r smol-grid, dev.args=list(bg="#222222")}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| fig-align: center
smol_grid <- long_grid(x = 1:20, y = 1:20)
ggplot(smol_grid) +
  geom_point(aes(x, y), colour = "white") + 
  theme_void() + 
  coord_equal()
```

Now let's compute the value of the simplex noise pattern at each of these points using `gen_simplex()`, and represent that as the size of the plot marker (because I can't resist the urge to make something pretty), or more conventionally as a contour plot illustrating the "height" of the pattern at each point:

```{r smol-simplex, dev.args=list(bg="#222222")}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| fig-align: center
#| column: screen-inset
#| layout-nrow: 1

smol_simplex <- smol_grid |>
  mutate(z = gen_simplex(x, y, seed = 1, frequency = .1)) 

smol_simplex |>
  ggplot(aes(x, y, size = z)) +
  geom_point(colour = "white", show.legend = FALSE) + 
  theme_void() + 
  coord_equal()
smol_simplex |>
  ggplot(aes(x, y, z = z)) +
  geom_contour_filled(show.legend = FALSE, bins = 10) + 
  theme_void() + 
  coord_equal()
```

Now imagine placing a ball at a point on this surface. Unless it's at a completely flat spot, it will start rolling in a particular direction and at a particular speed. We can work out where it will start rolling by computing the slope of surface at each point. To do this, we'll use a finite differencing approximation to calculate the partial derivatives. Or, to put it in less fancy terms, we'll add a small diffrenece `eps` to the x-coordinate and compute the value of the simplex noise at the modified values. That gives us the slope in the x-direction. We do the same thing for the y-direction. Putting these two vectors together gives us the local slope. 

```{r smol-diff, dev.args=list(bg="#222222")}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| fig-align: center
eps <- .001
smol_curl <- smol_grid |> mutate(
  x_add = gen_simplex(x + eps, y, seed = 1, frequency = .1),
  x_sub = gen_simplex(x - eps, y, seed = 1, frequency = .1),
  y_add = gen_simplex(x, y + eps, seed = 1, frequency = .1),
  y_sub = gen_simplex(x, y - eps, seed = 1, frequency = .1),
  x_slope = (x_add - x_sub) / (2 * eps), 
  y_slope = (y_add - y_sub) / (2 * eps),
  x_curl = -y_slope, 
  y_curl = x_slope
)
```

If I wanted to plot how fast the simplex noise field was changing at each point on this grid, I'd just plot the `x_slope` and `y_slope` values. That would give me something like this:

```{r smol-slope, dev.args=list(bg="#222222")}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| fig-align: center
ggplot(smol_curl) + 
  geom_segment(
    mapping = aes(
      x = x, 
      y = y, 
      xend = x + x_slope * 2, 
      yend = y + y_slope * 2
    ), 
    colour = "white", 
    arrow = arrow(length = unit(0.1, "cm"))
  ) + 
  theme_void() + 
  coord_equal()
```

This map of arrows (a.k.a. vector field) depicts the slope at each point on the simplex noise surface. It's a measure of how fast a ball would start rolling if you placed it down at a particular spot.

Let's tweak the analogy slightly. Instead of a ball on a hill, imagine the arrows depict a current pushing a rough-edged disc around in a pool of water. When we place the disc in the water it will start moving because the current pushes it around, but because it's rough-edged the friction of water flowing over it will make it start rotating. The *curl* of a field describes these rotational forces. In the same way that our simplex noise pattern implies a vector field of slope values, it also implies a vector field of curl values. For reasons that I'm sure a physicist can explain to me -- that I'm certain will have something to do with a conservation law of some kind -- `x_curl = -y_slope` and `y_curl = x_slope`. 

Whatever.

Anyway.

Now we have the curl of our simplex noise and we know vaguely what it means. More importantly, we can draw a pretty picture of the curl field:

```{r smol-curl, dev.args=list(bg="#222222")}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| fig-align: center
ggplot(smol_curl) + 
  geom_segment(
    mapping = aes(
      x = x, 
      y = y, 
      xend = x + x_curl * 2, 
      yend = y + y_curl * 2
    ), 
    colour = "white", 
    arrow = arrow(length = unit(0.1, "cm"))
  ) + 
  theme_void() + 
  coord_equal()
```

As it turns out I actually didn't need to bother with computing this manually, because `ambient` supplies a `curl_noise()` function that does the exact same computations for us. I pass it the `x` and `y` coordinates from my `smol_grid`, specify that the `generator` function is `gen_simplex()`, and pass the parameters of the noise function (e.g., its `seed` and `frequency`) as additional arguments:

```{r smol-curl2-make, dev.args=list(bg="#222222")}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| fig-align: center
curl <- curl_noise(
  generator = gen_simplex,
  seed = 1,
  frequency = .1,
  x = smol_grid$x, 
  y = smol_grid$y
)

as_tibble(curl)
```

This `curl` data frame contains `x` and `y` columns that specify the curl values at each point in the input. So now I can plot these curl values in the same "map of arrows" style, and unsurprisingly I obtain the same result as last time: 

```{r smol-curl2, dev.args=list(bg="#222222")}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| fig-align: center
smol_grid |>
  mutate(
    x2 = x + curl$x * 2,
    y2 = y + curl$y * 2
  ) |> 
  ggplot() + 
  geom_segment(
    mapping = aes(x, y, xend = x2, yend = y2),
    colour = "white", 
    arrow = arrow(length = unit(0.1, "cm"))
  ) + 
  theme_void() + 
  coord_equal()
```

Generative artists have a particular fondness for computing the curl of a noise field and using it for nefarious purposes. There are technical reasons for that, no doubt, but I'm lazy and I feel like I've spent too much of my life thinking about this already. So let's skip the reasons this time and just start doing it. To make my life a little easier I'll write an `update_curl()` function that takes a `current_state` data frame as input (which we assume contains variables `x` and `y` that define a grid), computes the curl at all points in this grid, and then returns a new set of `x` and `y` values that "add a little bit of curl" to those `x` and `y` values:

```{r update-curl}
update_curl <- function(current_state, step_size = .0005, ...) {
  curl <- curl_noise(
    x = current_state$x, 
    y = current_state$y,
    ...
  )
  next_state <- current_state |>
    mutate(
      x = x + curl$x * step_size,
      y = y + curl$y * step_size,
      time = time + 1
    )
  return(next_state)
}
```

Next, let's define an initial state. At "time" point 1 we have a set of co-ordinates laid out on a grid:

```{r initial-curl}
coords <- seq(0, 1, length.out = 50)
time_1 <- long_grid(x = coords, y = coords) |> 
  mutate(id = row_number(), time = 1)
time_1
```

Now we can use our `update_curl()` function to compute a new set of `x` and `y` values. We can do this multiple times if we like:

```{r evolve-curl, dev.args=list(bg="#222222")}
time_2 <- time_1 |>
  update_curl(
    generator = gen_simplex,
    frequency = 10, 
    seed = 1234
  )

time_3 <- time_2 |> 
  update_curl(
    generator = gen_simplex,
    frequency = 10, 
    seed = 1234
  )

time_3
```

At this point it's important to notice something of particular relevance to generative art. Are there any physicists near you as you read this? Can you hear them sighing? 

Good.

From a physics perspective I've done something quite peculiar in this code. I've updated the "position" of a set of points (or particles) by adding their *rotation* (i.e. curl) to their current "position". I'm not really simulating real movement in physical space I'm plotting changes in rotational forces. Curl fields don't plot real world movement, they're an abstraction.

Which is fine. From an artistic point of view we care mostly about the fact that we can use this tool to make pretty things. So let's visualise these "curl updates":

```{r simple-curl-noise, dev.args=list(bg="#222222")}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| fig-align: center
#| column: screen-inset
#| layout-nrow: 1
dat12 <- bind_rows(time_1, time_2)
dat123 <- bind_rows(time_1, time_2, time_3)

dat12 |>
  ggplot(aes(x, y, group = id)) + 
  geom_path(colour = "white") +
  theme_void() + 
  coord_equal() 
dat123 |>
  ggplot(aes(x, y, group = id)) + 
  geom_path(colour = "white") +
  theme_void() + 
  coord_equal() 

```

This seems promising, right?

## Curl of a fractal pattern

One nice thing about `curl_noise()` is that it can be applied to any generator, including `fracture()`. Here's the basic idea:

```{r curl-art-function}
curl_data <- function(
    data, 
    iterations = 50,
    step_size = .001,
    ...
) {
  
  update <- function(current_state, iteration, ...) {
    curl <- curl_noise(
      x = current_state$x, 
      y = current_state$y,
      generator = fracture,
      ...
    )
    next_state <- current_state |>
      mutate(
        x = x + curl$x * step_size,
        y = y + curl$y * step_size,
        time = time + 1
      )
    return(next_state)
  }
  
  data |> 
    mutate(id = row_number(), time = 1) |>
    accumulate(1:iterations, update, .init = _, ...) |>
    bind_rows()
}

curl_art <- function(...) {
  curl_data(...) |> 
    ggplot(aes(x, y, group = id)) + 
    geom_path(colour = "white") +
    theme_void() + 
    coord_equal() 
}
```

A grid of small fractal walks:

```{r curl-art-examples, dev.args=list(bg="#222222")}
#| fig-width: 10
#| fig-height: 10
#| fig-dpi: 300
#| fig-align: center
#| cache: true

smol_grid |>
  mutate(x = normalise(x), y = normalise(y)) |>
  curl_art(noise = gen_simplex, fractal = fbm, octaves = 4, freq_init = .5)
```


An example where the initial points all lie on a circle:

```{r curl-art-octaves, dev.args=list(bg="#222222")}
#| fig-width: 6
#| fig-height: 6
#| fig-dpi: 300
#| column: screen-inset
#| layout-nrow: 1
#| cache: true
circle <- function(n = 100) {
  tibble(
    theta = 2 * pi * (1:n) / n, 
    x = cos(theta),
    y = sin(theta)
  )
}

curl_circle <- function(octaves) {
  curl_art(
    data = circle(500),
    iterations = 100, 
    noise = gen_simplex,
    fractal = fbm,
    octaves = octaves, 
    seed = 1, 
    freq_init = 1,
    frequency = ~ . * 1.2,
    gain_init = 1,
    gain = ~ . * .9,
    step_size = .003
  )
}

curl_circle(octaves = 1)
curl_circle(octaves = 3)
curl_circle(octaves = 8)
```

A related example using polygons, heavily influenced by Thomas Lin Pedersen's "genesis" system:


```{r curl-art-examples-2, dev.args=list(bg="#222222")}
#| fig-width: 8
#| fig-height: 8
#| fig-dpi: 300
#| fig-align: center
#| cache: true

custom_curl_data <- function(data) {
  curl_data(
    data = data,
    iterations = 100, 
    octaves = 10,
    fractal = ridged,
    noise = gen_cubic,
    freq_init = 1,
    frequency = ~ . * 1.2,
    gain_init = 1,
    gain = ~ . * .9,
    seed = 1
  )
}

dat1 <- circle(5000) |> 
  custom_curl_data()

dat2 <- circle(5000) |>
  mutate(x = x * .99, y = y * .99) |>
  custom_curl_data()

ggplot(mapping = aes(x, y, group = time)) +
  geom_polygon(data = dat1, fill = "#ffffff10") +
  geom_polygon(data = dat2, fill = "#22222205") +
  theme_void() +
  coord_equal()
```


::: {.callout-important icon=false #exercise-curl-art}
## Exercises

The `curl-art-1.R` and `curl-art-2.R` scripts contain code to generate the "small grid of fractal walks" image and the "genesis-inspired" image. In both cases the ouput is written to a 2000x2000 pixel png file, and the time taken to complete the task printed to the screen.

1. Run both scripts, and compare the difference in rendering times.

2. Modify the "small grid" version so that it produces smoother looking results.

3. Explore what you can do with the "genesis style". It's a powerful technique that can do a lot more than I've done in this code. By now you should have a good sense of what dials you can turn: the fractal, the generator, the parameters, the palette, etc. You can use dplyr to modify the data if you want to. Try to make something you really like!

:::

